---
title: "Types of Data"
---

# Types of Data {#sec-types-of-data}

```{r}
#| label: setup
#| include: false

if (!require("pak")) install.packages("pak")
pak::pak(c("tidyverse", "knitr", "kableExtra"), ask = FALSE)

library(tidyverse)
library(knitr)
library(kableExtra)
```

## Overview

Broadly, data are either **numerical** or **categorical**. Other terms to describe this are quantitative or qualitative.

## Categorical (Qualitative) Data

Categorical data tells us which category an individual belongs to.

### Nominal Scale

Categories are distinguished by a **name**, with **no intrinsic ordering**.

**Examples:** sex, histology, cancer type, postcode

### Ordinal Scale

Categories are distinguished by name, **with intrinsic ordering**.

**Examples:** performance status, toxicity grade, tumour stage

### Dichotomous Variables

A categorical measure with only two categories (for example alive or dead) is called **dichotomous**. Sometimes the categories of a dichotomous variable are labelled 0 and 1, and are called a **binary variable**.

## Numerical (Quantitative) Data

Numerical data consists of values that are counts or measurements.

### Discrete Data

Values arise from a **counting** process.

**Examples:** number of tumours, number of metastases, number of hospital admissions

### Continuous Data

Values arise from a **measuring** process.

**Examples:** height, tumour size, planning treatment volume, age, overall survival, weight, FEV1

```{r}
#| label: fig-data-types
#| fig-cap: "Classification of data types"
#| fig-width: 10
#| fig-height: 6

library(ggplot2)

# Create diagram using ggplot
ggplot() +
  # Main categorical box
  annotate("rect", xmin = 0, xmax = 4, ymin = 4, ymax = 5, 
           fill = "#3498db", alpha = 0.3, colour = "#2980b9") +
  annotate("text", x = 2, y = 4.7, label = "Categorical (qualitative)", 
           fontface = "bold", size = 4) +
  annotate("text", x = 2, y = 4.3, label = "tells us which category an\nindividual belongs to", 
           size = 3) +
  
  # Nominal box
  annotate("rect", xmin = 0, xmax = 1.9, ymin = 2, ymax = 3.5, 
           fill = "#3498db", alpha = 0.2, colour = "#2980b9") +
  annotate("text", x = 0.95, y = 3.2, label = "Nominal scale", 
           fontface = "bold", size = 3.5) +
  annotate("text", x = 0.95, y = 2.8, label = "Categories distinguished\nby name, with no\nintrinsic ordering", 
           size = 2.5) +
  annotate("text", x = 0.95, y = 2.2, label = "(e.g. sex, histology,\ncancer type)", 
           size = 2.5, fontface = "italic") +
  
  # Ordinal box
  annotate("rect", xmin = 2.1, xmax = 4, ymin = 2, ymax = 3.5, 
           fill = "#3498db", alpha = 0.2, colour = "#2980b9") +
  annotate("text", x = 3.05, y = 3.2, label = "Ordinal scale", 
           fontface = "bold", size = 3.5) +
  annotate("text", x = 3.05, y = 2.8, label = "Categories distinguished\nby name, with\nintrinsic ordering", 
           size = 2.5) +
  annotate("text", x = 3.05, y = 2.2, label = "(e.g. performance status,\ntoxicity grade)", 
           size = 2.5, fontface = "italic") +
  
  # Main numerical box
  annotate("rect", xmin = 5, xmax = 9, ymin = 4, ymax = 5, 
           fill = "#e74c3c", alpha = 0.3, colour = "#c0392b") +
  annotate("text", x = 7, y = 4.7, label = "Numerical (quantitative)", 
           fontface = "bold", size = 4) +
  annotate("text", x = 7, y = 4.3, label = "values are counts or\nmeasurements", 
           size = 3) +
  
  # Discrete box
  annotate("rect", xmin = 5, xmax = 6.9, ymin = 2, ymax = 3.5, 
           fill = "#e74c3c", alpha = 0.2, colour = "#c0392b") +
  annotate("text", x = 5.95, y = 3.2, label = "Discrete", 
           fontface = "bold", size = 3.5) +
  annotate("text", x = 5.95, y = 2.8, label = "Values arise from\ncounting process", 
           size = 2.5) +
  annotate("text", x = 5.95, y = 2.2, label = "(e.g. number of\ntumours)", 
           size = 2.5, fontface = "italic") +
  
  # Continuous box
  annotate("rect", xmin = 7.1, xmax = 9, ymin = 2, ymax = 3.5, 
           fill = "#e74c3c", alpha = 0.2, colour = "#c0392b") +
  annotate("text", x = 8.05, y = 3.2, label = "Continuous", 
           fontface = "bold", size = 3.5) +
  annotate("text", x = 8.05, y = 2.8, label = "Values arise from\nmeasuring process", 
           size = 2.5) +
  annotate("text", x = 8.05, y = 2.2, label = "(e.g. height, tumour size,\nage, survival)", 
           size = 2.5, fontface = "italic") +
  
  # Connecting lines
  annotate("segment", x = 1, y = 4, xend = 1, yend = 3.5) +
  annotate("segment", x = 3, y = 4, xend = 3, yend = 3.5) +
  annotate("segment", x = 1, y = 4, xend = 3, yend = 4) +
  annotate("segment", x = 2, y = 4.0, xend = 2, yend = 4) +
  
  annotate("segment", x = 6, y = 4, xend = 6, yend = 3.5) +
  annotate("segment", x = 8, y = 4, xend = 8, yend = 3.5) +
  annotate("segment", x = 6, y = 4, xend = 8, yend = 4) +
  annotate("segment", x = 7, y = 4.0, xend = 7, yend = 4) +
  
  theme_void() +
  coord_cartesian(xlim = c(-0.5, 9.5), ylim = c(1.5, 5.5))
```

## Paired Data

When the same variable is measured on two occasions in the same individual, this is called **paired** data. If measurements are only made once on each individual they are **unpaired**.

Paired observations occur, for example, when a patient is assessed before and after a treatment (e.g. blood pressure before and after starting an anti-hypertensive), or when two sites on the same patient are compared (e.g. left and right eyes).

```{r}
#| label: tbl-paired-example
#| tbl-cap: "Example of paired data: blood pressure before and after treatment"

tibble(
  Patient = 1:6,
  `BP Before (mmHg)` = c(145, 160, 138, 155, 142, 168),
  `BP After (mmHg)` = c(130, 145, 132, 140, 135, 150)
) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Summary

| Data Type | Subtype | Description | Examples |
|-----------|---------|-------------|----------|
| Categorical | Nominal | Categories with no ordering | Sex, histology, cancer type |
| Categorical | Ordinal | Categories with ordering | Performance status, toxicity grade |
| Categorical | Dichotomous | Only two categories | Alive/dead, yes/no |
| Numerical | Discrete | Counting process | Number of tumours |
| Numerical | Continuous | Measuring process | Height, tumour size, survival time |
