---
title: "Types of Data"
---


```{r}
#| label: setup
#| include: false

if (!require("pak")) install.packages("pak")
pak::pak(c("tidyverse", "knitr", "kableExtra"), ask = FALSE)

library(tidyverse)
library(knitr)
library(kableExtra)
```

## Overview

Broadly, data are either **numerical** or **categorical**. Other terms to describe this are quantitative or qualitative.

## Categorical (Qualitative) Data

Categorical data tells us which category an individual belongs to.

### Nominal Scale

Categories are distinguished by a **name**, with **no intrinsic ordering**.

**Examples:** sex, histology, cancer type, postcode

### Ordinal Scale

Categories are distinguished by name, **with intrinsic ordering**.

**Examples:** performance status, toxicity grade, tumour stage

### Dichotomous Variables

A categorical measure with only two categories (for example alive or dead) is called **dichotomous**. Sometimes the categories of a dichotomous variable are labelled 0 and 1, and are called a **binary variable**.

## Numerical (Quantitative) Data

Numerical data consists of values that are counts or measurements.

### Discrete Data

Values arise from a **counting** process.

**Examples:** number of tumours, number of metastases, number of hospital admissions

### Continuous Data

Values arise from a **measuring** process.

**Examples:** height, tumour size, planning treatment volume, age, overall survival, weight, FEV1

```{r}
#| label: fig-data-types
#| fig-cap: "Classification of data types"
#| fig-width: 10
#| fig-height: 6

# Create diagram using ggplot
ggplot() +
  # Main categorical box
  annotate("rect", xmin = 0, xmax = 4, ymin = 4, ymax = 5, 
           fill = "#3498db", alpha = 0.3, colour = "#2980b9") +
  annotate("text", x = 2, y = 4.7, label = "Categorical (qualitative)", 
           fontface = "bold", size = 4) +
  annotate("text", x = 2, y = 4.3, label = "tells us which category an\nindividual belongs to", 
           size = 3) +
  
  # Nominal box
  annotate("rect", xmin = 0, xmax = 1.9, ymin = 2, ymax = 3.5, 
           fill = "#3498db", alpha = 0.2, colour = "#2980b9") +
  annotate("text", x = 0.95, y = 3.2, label = "Nominal scale", 
           fontface = "bold", size = 3.5) +
  annotate("text", x = 0.95, y = 2.8, label = "Categories distinguished\nby name, with no\nintrinsic ordering", 
           size = 2.5) +
  annotate("text", x = 0.95, y = 2.2, label = "(e.g. sex, histology,\ncancer type)", 
           size = 2.5, fontface = "italic") +
  
  # Ordinal box
  annotate("rect", xmin = 2.1, xmax = 4, ymin = 2, ymax = 3.5, 
           fill = "#3498db", alpha = 0.2, colour = "#2980b9") +
  annotate("text", x = 3.05, y = 3.2, label = "Ordinal scale", 
           fontface = "bold", size = 3.5) +
  annotate("text", x = 3.05, y = 2.8, label = "Categories distinguished\nby name, with\nintrinsic ordering", 
           size = 2.5) +
  annotate("text", x = 3.05, y = 2.2, label = "(e.g. performance status,\ntoxicity grade)", 
           size = 2.5, fontface = "italic") +
  
  # Main numerical box
  annotate("rect", xmin = 5, xmax = 9, ymin = 4, ymax = 5, 
           fill = "#e74c3c", alpha = 0.3, colour = "#c0392b") +
  annotate("text", x = 7, y = 4.7, label = "Numerical (quantitative)", 
           fontface = "bold", size = 4) +
  annotate("text", x = 7, y = 4.3, label = "values are counts or\nmeasurements", 
           size = 3) +
  
  # Discrete box
  annotate("rect", xmin = 5, xmax = 6.9, ymin = 2, ymax = 3.5, 
           fill = "#e74c3c", alpha = 0.2, colour = "#c0392b") +
  annotate("text", x = 5.95, y = 3.2, label = "Discrete", 
           fontface = "bold", size = 3.5) +
  annotate("text", x = 5.95, y = 2.8, label = "Values arise from\ncounting process", 
           size = 2.5) +
  annotate("text", x = 5.95, y = 2.2, label = "(e.g. number of\ntumours)", 
           size = 2.5, fontface = "italic") +
  
  # Continuous box
  annotate("rect", xmin = 7.1, xmax = 9, ymin = 2, ymax = 3.5, 
           fill = "#e74c3c", alpha = 0.2, colour = "#c0392b") +
  annotate("text", x = 8.05, y = 3.2, label = "Continuous", 
           fontface = "bold", size = 3.5) +
  annotate("text", x = 8.05, y = 2.8, label = "Values arise from\nmeasuring process", 
           size = 2.5) +
  annotate("text", x = 8.05, y = 2.2, label = "(e.g. height, tumour size,\nage, survival)", 
           size = 2.5, fontface = "italic") +
  
  # Connecting lines
  annotate("segment", x = 1, y = 4, xend = 1, yend = 3.5) +
  annotate("segment", x = 3, y = 4, xend = 3, yend = 3.5) +
  annotate("segment", x = 1, y = 4, xend = 3, yend = 4) +
  annotate("segment", x = 2, y = 4.0, xend = 2, yend = 4) +
  
  annotate("segment", x = 6, y = 4, xend = 6, yend = 3.5) +
  annotate("segment", x = 8, y = 4, xend = 8, yend = 3.5) +
  annotate("segment", x = 6, y = 4, xend = 8, yend = 4) +
  annotate("segment", x = 7, y = 4.0, xend = 7, yend = 4) +
  
  theme_void() +
  coord_cartesian(xlim = c(-0.5, 9.5), ylim = c(1.5, 5.5))
```

## Paired Data

The majority of statistical analyses compare characteristics measured in two separate groups of individuals. In some circumstances, however, data may consist of **pairs of outcome measurements**.

When the same variable is measured on two occasions in the same individual, this is called **paired** data. If measurements are only made once on each individual they are **unpaired**.

### Examples of Paired Data

**Before and after treatment:** We might wish, for example, to carry out a study where the assessment of tumour response to radiotherapy is based on comparing tumour size measurements in a group of lung cancer patients, before and after they received treatment. For each person, we therefore have a pair of measures: tumour size after treatment and tumour size before treatment.

**Comparing two sites:** When two measurements are taken on the same patient (e.g., comparing left and right eyes, or two different anatomical sites).

::: {.callout-important}
## Why Pairing Matters
It is important to take this pairing in the data into account when assessing how much on average the treatment has affected tumour size. Paired analyses account for within-person variability and are typically more powerful than unpaired analyses.
:::

```{r}
#| label: tbl-paired-example
#| tbl-cap: "Example of paired data: tumour size before and after radiotherapy in lung cancer patients"

tibble(
  Patient = 1:8,
  `Tumour Size Before (cm)` = c(4.2, 3.8, 5.1, 4.5, 3.2, 6.0, 4.8, 3.5),
  `Tumour Size After (cm)` = c(2.1, 2.5, 3.2, 2.8, 1.8, 4.2, 3.1, 2.0),
  `Change (cm)` = c(-2.1, -1.3, -1.9, -1.7, -1.4, -1.8, -1.7, -1.5)
) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Summary

| Data Type | Subtype | Description | Examples |
|-----------|---------|-------------|----------|
| Categorical | Nominal | Categories with no ordering | Sex, histology, cancer type |
| Categorical | Ordinal | Categories with ordering | Performance status, toxicity grade |
| Categorical | Dichotomous | Only two categories | Alive/dead, yes/no |
| Numerical | Discrete | Counting process | Number of tumours |
| Numerical | Continuous | Measuring process | Height, tumour size, survival time |
