---
title: "Survival Analysis"
---


```{r}
#| label: setup
#| include: false

if (!require("pak")) install.packages("pak")
pak::pak(c("tidyverse", "survival", "survminer", "knitr", "kableExtra"), ask = FALSE)

library(tidyverse)
library(survival)
library(survminer)
library(knitr)
library(kableExtra)
```

## Introduction

**Survival analysis** involves the use of data which measures the time to an 'event'. In the context of cancer, the event could be death and the survival data might record time from cancer diagnosis to death.

Other survival events of interest can include:

- Time until disease recurrence
- Time to re-hospitalisation following discharge
- Progression-free survival
- Time to treatment failure

Survival data are concerned with recording the **length of time** for a patient to reach the specific endpoint, rather than simply recording whether the endpoint was reached. It therefore involves **two variables** – time to the event and whether the event occurred.

## Censoring

One problem with survival data is that there is often incomplete follow-up for patients. This is known as **censoring**.

Censoring arises when:

- A study is finished before all patients experience the event (which could be death)
- Patients have to be excluded from the study due to other reasons (migration, lost to follow-up, other adverse event)

```{r}
#| label: fig-censoring
#| fig-cap: "Schematic of a survival analysis study. Each line represents a subject. Circles indicate censored data, crosses indicate the event occurred."
#| fig-width: 8
#| fig-height: 5

# Create example survival data for visualization
set.seed(123)
survival_schematic <- tibble(
  patient = factor(1:8),
  start = 0,
  end = c(95, 78, 25, 100, 55, 30, 85, 70),
  status = c("Censored", "Censored", "Event", "Censored", "Event", "Event", "Event", "Censored")
)

ggplot(survival_schematic, aes(y = patient)) +
  geom_segment(aes(x = start, xend = end, yend = patient), linewidth = 1, colour = "#3498db") +
  geom_point(aes(x = end, shape = status, colour = status), size = 4) +
  scale_shape_manual(values = c("Censored" = 1, "Event" = 4)) +
  scale_colour_manual(values = c("Censored" = "#3498db", "Event" = "#e74c3c")) +
  labs(x = "Survival time (months)", y = "Patient", 
       shape = "Status", colour = "Status") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "top")
```

A comparison of the mean time to reach the endpoint in patients can give misleading results because of censored data. Therefore, a number of statistical techniques, known as **survival methods**, have been developed to deal with these situations.

The aim of these techniques is to:

- Statistically describe survival times
- Compare survival over several therapy groups (the idea being the longer the survival times, the better the therapy)
- Find relations between survival times and other prognostic variables (for example, age, stage, severity of disease)

## Kaplan-Meier Survival Curves

Survival data can be displayed graphically in a **Kaplan-Meier plot**.

- The **vertical axis** displays the proportion of patients remaining free of the endpoint at any time after baseline
- Each **event** (death) is indicated by a step in the curve
- **Censored data** are indicated by (+)
- The resulting curve is therefore a series of steps, starting at a survival probability of 1 (or 100%) at time 0 and drops towards 0 as time increases

In effect it shows the proportion of people who are still at risk of experiencing the event, while the horizontal axis shows the time that the subjects were followed up for.

The plot can be used to:

- Describe time-to-event in one group
- Compare time-to-event among different groups (for example, by treatment, or sex)

```{r}
#| label: fig-km-curve
#| fig-cap: "Kaplan-Meier curves for overall survival among renal cancer patients with brain metastases treated with different radiotherapy modalities"
#| fig-width: 9
#| fig-height: 6

# Create synthetic survival data similar to the brain metastases example
set.seed(456)

# WBRT - worst survival (median ~60 days)
wbrt <- tibble(
  time = c(rexp(10, 1/60), runif(4, 30, 180)),
  status = c(rep(1, 10), rep(0, 4)),
  treatment = "WBRT"
)

# SRS - better survival (median ~426 days)
srs <- tibble(
  time = c(rexp(6, 1/400), runif(4, 200, 800)),
  status = c(rep(1, 6), rep(0, 4)),
  treatment = "SRS"
)

# PBRT - best survival (median ~588 days)
pbrt <- tibble(
  time = c(rexp(4, 1/550), runif(4, 300, 1000)),
  status = c(rep(1, 4), rep(0, 4)),
  treatment = "PBRT"
)

surv_data <- bind_rows(wbrt, srs, pbrt)
surv_data$treatment <- factor(surv_data$treatment, levels = c("WBRT", "SRS", "PBRT"))

# Convert time to years
surv_data$time_years <- surv_data$time / 365

# Fit survival curves
fit <- survfit(Surv(time_years, status) ~ treatment, data = surv_data)

# Plot
ggsurvplot(fit, data = surv_data,
           palette = c("#e74c3c", "#3498db", "#27ae60"),
           risk.table = TRUE,
           pval = TRUE,
           conf.int = FALSE,
           xlab = "Time since end of radiotherapy (years)",
           ylab = "Percent alive (%)",
           legend.title = "Treatment",
           legend.labs = c("WBRT", "SRS", "PBRT"),
           surv.scale = "percent",
           break.time.by = 1,
           ggtheme = theme_minimal(base_size = 12))
```

### Interpreting Kaplan-Meier Curves

At the start of the study, no patients have died and the proportion at risk is **1.0 (100%)**. At each time point where a line drops, at least 1 patient experiences the event (death).

The dashed horizontal line at 50% indicates the **median survival time** in each treatment group. The median survival time is a useful summary measure that indicates the time at which 50% of patients have experienced the event.

## Log-Rank Test

The most common (non-parametric) method of comparing survival between independent groups is the **log-rank test**.

The null hypothesis would be that the groups have **equal median survival times** (i.e. there is no difference in median survival between the groups). The alternative hypothesis is that at least one of the groups is different.

If the p-value is less than 0.05, the null hypothesis is rejected and we can conclude that there is evidence of at least one difference in the median survival times.

### Extensions to the Log-Rank Test

- **Stratified log-rank test**: Can adjust for categorical prognostic covariates (for example age group or sex)
- **Log-rank test for trend**: Used for ordered groups (for example, cancer stage, or number of metastases). This is a more appropriate test for considering a trend in survival across the groups

## Life Tables

**Life tables** describe survival data, where the results have been grouped into time intervals, often of equal length. This method is often described as **actuarial**.

The method of calculation is similar to the Kaplan-Meier method, but differences arise because of the lack of precision of recording of survival times. In general, the **Kaplan-Meier analysis is recommended**.

## Cox Proportional Hazards Model

The log-rank test is solely a hypothesis test, comparing survival in two or more groups. It does not allow the relationships between one or more categorical or continuous factors and survival to be quantified.

To do this we employ the **hazard rate** (sometimes called the failure rate) which is closely related to survival curves. The hazard rate represents the risk of dying in a very short time interval after a given time t, assuming survival to time t.

We want to know whether there are any systematic differences between the hazards, over all time points, of individuals with different characteristics.

This can be achieved using the **Cox proportional hazards model** to test the independent effects of a number of explanatory variables on the hazard. The method assumes that the effect of a predictor on survival **does not change over time**.

### Output from Cox Models

For each variable the Cox proportional hazards model produces an output which includes:

- A **hazard ratio** with a confidence interval
- A **p-value**

### Interpreting Hazard Ratios

| Hazard Ratio | Interpretation |
|--------------|----------------|
| **> 1** | Raised hazard, suggests **shorter** survival times. The hazard increases by 100 × (HR - 1)% |
| **= 1** | No increased or decreased hazard of the endpoint |
| **< 1** | Decreased hazard, suggests **longer** survival times. The hazard decreases by 100 × (1 - HR)% |

The **p-value** relates to a significance test performed to test whether that hazard ratio is different from one.

### Types of Predictors

- For a **continuous predictor**: The hazard ratio describes what happens to the hazard as the predictor changes by 1 unit of measurement
- For levels of a **categorical predictor**: The hazard ratio describes the relative difference in the hazard between a level and a chosen reference level

## Worked Example: Brain Metastases in Renal Cancer

Using data from the brain metastases example, a proportional hazards model produces:

```{r}
#| label: tbl-cox-output
#| tbl-cap: "Cox proportional hazards model output for brain metastases survival"

cox_output <- tibble(
  Variable = c("Sex", "  Men (reference)", "  Women", 
               "Age (years)", 
               "Number of metastases", "  1 (reference)", "  2+",
               "Treatment", "  SRS (reference)", "  WBRT", "  PBRT"),
  `Hazard Ratio` = c("", "1", "1.33", "1.02", "", "1", "1.68", "", "1", "7.11", "0.75"),
  `95% CI` = c("", "-", "0.24-2.36", "0.97-1.08", "", "-", "0.45-6.19", "", "-", "1.85-27.3", "0.23-2.36"),
  `P-value` = c("", "-", "0.54", "0.40", "", "-", "0.43", "", "-", "0.004", "0.62")
)

cox_output |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover"))
```

### Interpretation

**Sex:** The reference group is men. After accounting for age, treatment and number of brain metastases, the hazard for women is higher than the hazard for men by 33% (1.33 - 1). The p-value is greater than 0.05 and the confidence interval includes 1, so there is **no evidence** to suggest an association between sex and hazard in the wider population.

**Age:** Age is a continuous variable. After accounting for sex and treatment, increasing age is associated with increasing hazard. A one-year increase in age is associated with a 2% (1.02 - 1) increase in hazard. However, the p-value is above 0.05 and the confidence interval includes 1, so there is **no evidence** to suggest an association between age and hazard in the wider population.

**Treatment:** The treatment reference group is SRS.

- **WBRT:** After accounting for age and sex, the hazard for WBRT is statistically significantly higher than the hazard for SRS. Hazard for WBRT patients is **611% (7.11 - 1) higher** than for SRS. This increase is statistically significantly different from 0. In a wider population of similarly chosen patients, the increased hazard is likely to be between 85% and 2630%.
- **PBRT:** For patients chosen to receive PBRT, the hazard is 25% (1 - 0.75) lower than SRS, but is not significantly different from 0.

## Summary

| Concept | Description |
|---------|-------------|
| Survival analysis | Analysis of time-to-event data |
| Censoring | Incomplete follow-up (event not observed) |
| Kaplan-Meier curve | Graphical display of survival probability over time |
| Median survival | Time at which 50% of patients have experienced the event |
| Log-rank test | Non-parametric test comparing survival between groups |
| Cox proportional hazards | Regression model for survival data |
| Hazard ratio | Relative risk of event; HR > 1 = worse survival, HR < 1 = better survival |
